### 2024-04-02 output/paper.qmd
Draft Submission

---
title: "The Next Generation of Canadian Salmon"
subtitle: "The Effects of the Commercial Fishing Industry on Spawning Rates"
author: 
  - Timothius Prajogi
thanks: "Code and data are available at: https://github.com/prajogt/canadian_salmon_spawn.git ."
date: today
date-format: long
abstract: "Using the NuSEDS dataset provided by the Government of Canada and statistics published by Fisheries and Oceans Canada regarding the number of commercial fishing licences issued by Pacific Region, we investigate the effects of commercial fishing on the reproductive capacity of Canadian Salmon. In tandem, we investigate how much of the change in spawn rates can be attributed to the health of the ecosystem, climate, and the fish that live in it. This paper discusses the impact of direct human intervention in the form of commercial fishing, aiming to highlight the importance in proper management and control in the fishing industry in alignement with changes in the environment by predicting spawning trends of future generations."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
header-includes:
  \usepackage{placeins}
---

# Introduction

The declining trend in Canadian salmon runs, as observed by many avid anglers and supported by empirical data (INSERT CITATION), serves as the focal point of this paper. Once thriving waterways have experienced a significant reduction in salmon populations, posing challenges not only for recreational anglers but also for the fishing industry, a vital component of Canadian exports and numerous livelihoods, particularly in coastal regions. Due to this significance to many Canadians, it is important to investigate the decline in salmon runs, highlighting the importance of understanding and addressing these challenges that are faced in the effort to safeguard future salmon generations.

This paper will focus on commercial fishing trends, ocean climate trends, and spawning rates over the last decade, using datasets and fisheries statistics provided through the Government of Canada's open data portal. In particular, the NuSEDS (New Salmon Escapement Database System) provides key insights on spawner records and abundance estimates for freshwater streams and tributaries, with statistics provided by Fisheries and Oceans Canada providing detail on the amount of licenses issued and fish landed by species. Through this data we discover the potential drivers of salmon population decline, including but not limited to habitat degradation, climate change impacts, and human-induced factors such as over fishing and habitat loss. In identifying the critical areas for repair and reform, we aim to explore potential mitigation strategies and management plans with the goal of restoring salmon populations and promoting sustainable fishing practices, protecting this key resource. Preserving healthy salmon populations is not only essential for maintaining biodiversity and ecosystem balance but also for meeting demand required of this industry, both in labour and in product.

This paper will introduce the data and models we used to analyze these themes and trends, presenting our findings and their implications for future salmon generations.

# Data

# Model

# Results

# Discussion

# Appendix

# References

### 2024-04-17 scripts/01_download_data.R
download data

@@ -1,0 +1,31 @@
+#### Preamble ####
+# Purpose: Download Salmon Data
+# Author: Timothius Prajogi
+# Date: 18 April 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: none
+
+#### Workplace setup ####
+
+library(tidyverse)
+library(openxlsx)
+
+#### Download and save Salmon data ####
+
+# Get commercial catch data
+url <- "https://api-proxy.edh.azure.cloud.dfo-mpo.gc.ca/catalogue/records/7ac5fe02-308d-4fff-b805-80194f8ddeb4/attachments/ise-ecs.zip"
+
+download.file(url, "input/salmon-data-bundle.zip", mode = "wb")
+
+zip_file <- "input/salmon-data-bundle.zip"
+
+unzip(zip_file, exdir="input/")
+
+# Get salmon spawning data
+url <- "https://api-proxy.edh.azure.cloud.dfo-mpo.gc.ca/catalogue/records/c48669a3-045b-400d-b730-48aafe8c5ee6/attachments/All%20Areas_Simplified%20Version_20240110.xlsx"
+download.file(url, "input/nuseds.xlsx", mode = "wb")
+
+nuseds_data_xlsx <- read.xlsx("input/nuseds.xlsx")
+
+write_csv(nuseds_data_xlsx, "input/nuseds.csv")

### 2024-04-17 scripts/01_download_data.R
move data + add data dictionaries

@@ -16,16 +16,16 @@ library(openxlsx)
 # Get commercial catch data
 url <- "https://api-proxy.edh.azure.cloud.dfo-mpo.gc.ca/catalogue/records/7ac5fe02-308d-4fff-b805-80194f8ddeb4/attachments/ise-ecs.zip"
 
-download.file(url, "input/salmon-data-bundle.zip", mode = "wb")
+download.file(url, "input/data/salmon-data-bundle.zip", mode = "wb")
 
-zip_file <- "input/salmon-data-bundle.zip"
+zip_file <- "input/data/salmon-data-bundle.zip"
 
-unzip(zip_file, exdir="input/")
+unzip(zip_file, exdir="input/data/")
 
 # Get salmon spawning data
 url <- "https://api-proxy.edh.azure.cloud.dfo-mpo.gc.ca/catalogue/records/c48669a3-045b-400d-b730-48aafe8c5ee6/attachments/All%20Areas_Simplified%20Version_20240110.xlsx"
-download.file(url, "input/nuseds.xlsx", mode = "wb")
+download.file(url, "input/data/nuseds.xlsx", mode = "wb")
 
-nuseds_data_xlsx <- read.xlsx("input/nuseds.xlsx")
+nuseds_data_xlsx <- read.xlsx("input/data/nuseds.xlsx")
 
-write_csv(nuseds_data_xlsx, "input/nuseds.csv")
+write_csv(nuseds_data_xlsx, "input/data/nuseds.csv")

### 2024-04-17 scripts/02.1_clean_catch_data.R
Clean catch data

@@ -1,0 +1,96 @@
+#### Preamble ####
+# Purpose: Clean Salmon Catch Data
+# Author: Timothius Prajogi
+# Date: 18 April 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01_download_data
+
+#### Workplace setup ####
+
+library(tidyverse)
+library(arrow)
+
+#### Clean Salmon Catch Data ####
+
+# Read in data
+
+# Gill net
+gill_net_catches <- 
+  read_csv(
+    "input/data/ise-ecs/cs-gn-pac-dfo-mpo-science-eng.csv",
+    skip = 1
+  )
+
+# Seine
+seine_catches <-
+  read_csv(
+    "input/data/ise-ecs/cs-sn-pac-dfo-mpo-science-eng.csv",
+    skip = 1
+  )
+
+# Troll
+troll_catches <- 
+  read_csv(
+    "input/data/ise-ecs/cs-tr-pac-dfo-mpo-science-eng.csv",
+    skip = 1,
+    locale = locale(encoding = "latin1")
+  )
+
+# Replace all N/A values in catch and release values to be 0
+# (we add up all the values, therefore NA is the same as 0)
+# These values are values removed due to privacy restrictions, however
+# the inclusion of these rows are important for total vessel counts
+gill_net_catches <-
+  gill_net_catches |>
+  mutate_at(vars(starts_with(c(
+    "SOCKEYE_KEPT", "SOCKEYE_RELD",
+    "COHO_KEPT", "COHO_RELD",
+    "PINK_KEPT", "PINK_RELD",
+    "CHUM_KEPT", "CHUM_RELD",
+    "CHINOOK_KEPT", "CHINOOK_RELD",
+    "STEELHEAD_KEPT", "STEELHEAD_RELD"
+  ))), ~replace_na(., 0))
+
+seine_catches <-
+  seine_catches |>
+  mutate_at(vars(starts_with(c(
+    "SOCKEYE_KEPT", "SOCKEYE_RELD",
+    "COHO_KEPT", "COHO_RELD",
+    "PINK_KEPT", "PINK_RELD",
+    "CHUM_KEPT", "CHUM_RELD",
+    "CHINOOK_KEPT", "CHINOOK_RELD",
+    "STEELHEAD_KEPT", "STEELHEAD_RELD"
+  ))), ~replace_na(., 0))
+
+troll_catches <-
[…]

### 2024-04-17 scripts/01_download_data.R
Use full dataset

@@ -23,9 +23,17 @@ zip_file <- "input/data/salmon-data-bundle.zip"
 unzip(zip_file, exdir="input/data/")
 
 # Get salmon spawning data
-url <- "https://api-proxy.edh.azure.cloud.dfo-mpo.gc.ca/catalogue/records/c48669a3-045b-400d-b730-48aafe8c5ee6/attachments/All%20Areas_Simplified%20Version_20240110.xlsx"
+url <- "https://api-proxy.edh.azure.cloud.dfo-mpo.gc.ca/catalogue/records/c48669a3-045b-400d-b730-48aafe8c5ee6/attachments/All%20Areas%20NuSEDS_20240110.xlsx"
 download.file(url, "input/data/nuseds.xlsx", mode = "wb")
 
 nuseds_data_xlsx <- read.xlsx("input/data/nuseds.xlsx")
+
+# Filter only relevant columns, otherwise the file becomes too large to save to github
+nuseds_data_xlsx <- 
+  nuseds_data_xlsx |> 
+  select(-GAZETTED_NAME, -LOCAL_NAME_2, -ENUMERATION_METHODS, 
+         -NATURAL_ADULT_FEMALES, -NATURAL_ADULT_MALES, -EFFECTIVE_FEMALES, -WEIGHTED_PCT_SPAWN,
+         -WATERSHED_CDE, -POPULATION, -ACCURACY, -PRECISION, -INDEX_YN, -RELIABILITY, 
+         -ESTIMATE_STAGE, -ESTIMATE_CLASSIFICATION, -NO_INSPECTIONS_USED, -ESTIMATE_METHOD, -CREATED_DTT, -UPDATED_DTT)
 
 write_csv(nuseds_data_xlsx, "input/data/nuseds.csv")

### 2024-04-17 scripts/02.2_clean_spawn_data.R
Clean spawn data

@@ -1,0 +1,63 @@
+#### Preamble ####
+# Purpose: Clean Salmon Spawn Data
+# Author: Timothius Prajogi
+# Date: 18 April 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01_download_data
+
+#### Workplace setup ####
+
+library(tidyverse)
+library(arrow)
+
+#### Clean Salmon Spawn Data ####
+
+# Read in data
+
+# Gill net
+nuseds_data <- 
+  read_csv(
+    "input/data/nuseds.csv"
+  )
+
+# Rename cols
+nuseds_data <-
+  nuseds_data |> 
+  rename(YEAR = ANALYSIS_YR, LOCAL_NAME = LOCAL_NAME_1)
+
+# Split dataset to only include rows with specific features
+
+# Spawning population features
+spawning_population <-
+  nuseds_data |>
+  select(AREA, WATERBODY, LOCAL_NAME, YEAR, SPECIES,
+         NATURAL_ADULT_SPAWNERS, NATURAL_JACK_SPAWNERS, NATURAL_SPAWNERS_TOTAL, TOTAL_RETURN_TO_RIVER) |>
+  filter(!is.na(NATURAL_ADULT_SPAWNERS))
+  
+# Broodstock (amount taken out for the use of human influenced spawning)
+broodstock <-
+  nuseds_data |> 
+  select(AREA, WATERBODY, LOCAL_NAME, YEAR, SPECIES,
+         ADULT_BROODSTOCK_REMOVALS, JACK_BROODSTOCK_REMOVALS, TOTAL_BROODSTOCK_REMOVALS, OTHER_REMOVALS, TOTAL_RETURN_TO_RIVER) |>
+  filter(!is.na(ADULT_BROODSTOCK_REMOVALS) & !is.na(TOTAL_BROODSTOCK_REMOVALS) & !is.na(TOTAL_RETURN_TO_RIVER))
+
+# Spawn timing features (arrival, peak)
+spawn_timings <-
+  nuseds_data |> 
+  select(AREA, WATERBODY, LOCAL_NAME, YEAR, SPECIES,
+         STREAM_ARRIVAL_DT_FROM, STREAM_ARRIVAL_DT_TO, 
+         START_SPAWN_DT_FROM, START_SPAWN_DT_TO, 
+         PEAK_SPAWN_DT_FROM, PEAK_SPAWN_DT_TO, 
[…]

### 2024-04-17 scripts/03_aggregate_data.R
aggregated data

@@ -1,0 +1,168 @@
+#### Preamble ####
+# Purpose: Group together statistics based on species
+# Author: Timothius Prajogi
+# Date: 18 April 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01_download_data, 02.x_clean_data
+
+#### Workplace setup ####
+
+library(tidyverse)
+library(arrow)
+
+#### Aggregate By Species ####
+
+# Load in data
+gill_net_catches <- read_parquet("output/data/gill_net_catches.parquet")
+seine_catches <- read_parquet("output/data/seine_catches.parquet")
+troll_catches <- read_parquet("output/data/troll_catches.parquet")
+
+aggregated_gill_net_data <- 
+  gill_net_catches |> 
+  group_by(CALENDAR_YEAR, MGMT_AREA) |>
+  summarize(
+    VESSEL_COUNT = sum(VESSEL_COUNT, na.rm = TRUE),
+    BOAT_DAYS = sum(BOAT_DAYS, na.rm = TRUE),
+    SOCKEYE_KEPT = sum(SOCKEYE_KEPT, na.rm = TRUE),
+    SOCKEYE_RELD = sum(SOCKEYE_RELD, na.rm = TRUE),
+    COHO_KEPT = sum(COHO_KEPT, na.rm = TRUE),
+    COHO_RELD = sum(COHO_RELD, na.rm = TRUE),
+    PINK_KEPT = sum(PINK_KEPT, na.rm = TRUE),
+    PINK_RELD = sum(PINK_RELD, na.rm = TRUE),
+    CHUM_KEPT = sum(CHUM_KEPT, na.rm = TRUE),
+    CHUM_RELD = sum(CHUM_RELD, na.rm = TRUE),
+    CHINOOK_KEPT = sum(CHINOOK_KEPT, na.rm = TRUE),
+    CHINOOK_RELD = sum(CHINOOK_RELD, na.rm = TRUE),
+    STEELHEAD_KEPT = sum(STEELHEAD_KEPT, na.rm = TRUE),
+    STEELHEAD_RELD = sum(STEELHEAD_RELD, na.rm = TRUE)
+  )
+
+aggregated_seine_data <- 
+  seine_catches |> 
+  group_by(CALENDAR_YEAR, MGMT_AREA) |>
[…]

### 2024-04-17 scripts/03_aggregate_data.R
model population given catch data

@@ -128,11 +128,11 @@ pivoted_catch_data <-
   aggregated_catch_data |>
   pivot_longer(cols = -c(CALENDAR_YEAR, VESSEL_COUNT, BOAT_DAYS), names_to = "fish_action", values_to = "value") |>
   separate(fish_action, into = c("fish_species", "action"), sep = "_", remove = FALSE) |>
-  mutate(action = ifelse(action == "KEPT", "salmon_kept", "salmon_released")) |>
+  mutate(action = ifelse(action == "KEPT", "SALMON_KPT", "SALMON_RLD")) |>
   select(-fish_action) |>
   pivot_wider(names_from = action, values_from = value) |>
-  mutate(species = fish_species) |>
-  select(CALENDAR_YEAR, VESSEL_COUNT, BOAT_DAYS, species, salmon_kept, salmon_released)
+  mutate(SPECIES = fish_species) |>
+  select(CALENDAR_YEAR, VESSEL_COUNT, BOAT_DAYS, SPECIES, SALMON_KPT, SALMON_RLD)
 
 write_parquet(pivoted_catch_data, "output/data/aggregated_pivoted_catch_data.parquet")
 
@@ -151,6 +151,8 @@ aggregated_spawning_population_by_area <-
     TOTAL_RETURN_TO_RIVER = sum(TOTAL_RETURN_TO_RIVER, na.rm = TRUE),
     SALMON_POPULATION = sum(SALMON_POPULATION, na.rm = TRUE)
   )
+
+aggregated_spawning_population_by_area$SPECIES = toupper(aggregated_spawning_population_by_area$SPECIES)
 
 aggregated_spawning_population <- 
   aggregated_spawning_population_by_area |>

### 2024-04-17 scripts/04.1_catch_model.R

@@ -1,0 +1,46 @@
+#### Preamble ####
+# Purpose: Create LM predicting Salmon spawn using Catch rates
+# Author: Timothius Prajogi
+# Date: 18 April 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01_download_data, 02.x_clean_data, 03_aggregate_data
+
+#### Workplace setup ####
+
+library(rstanarm)
+library(tidyverse)
+library(modelsummary)
+
+set.seed(302)
+
+#### Create Model ####
+
+# Load in data
+catch_data <- read_parquet("output/data/aggregated_pivoted_catch_data.parquet") 
+
+catch_data <-
+  catch_data |>
+  rename(YEAR = CALENDAR_YEAR)
+
+spawning_population_data <- read_parquet("output/data/aggregated_spawning_population.parquet")
+
+# Since for this situation, the catch data only starts at 2002, only consider 
+# spawning data for those years
+spawning_population_data <-
+  spawning_population_data |>
+  filter(YEAR >= 2002) |>
+  select(YEAR, SPECIES, SALMON_POPULATION)
+
+model_data <- 
+  left_join(catch_data, spawning_population_data, by = c("YEAR", "SPECIES")) |>
+  mutate(SALMON_POPULATION = ifelse(is.na(SALMON_POPULATION), 0, SALMON_POPULATION)) |>
+  filter(SPECIES != "STEELHEAD") # Not common
+
+catch_population_model <- lm(SALMON_POPULATION ~ VESSEL_COUNT + BOAT_DAYS + SALMON_KPT + SALMON_RLD, data = model_data)
+
+saveRDS(
+  catch_population_model,
+  file = "output/models/catch_population_model.rds"
+)
+

### 2024-04-17 scripts/03_aggregate_data.R
add comments

@@ -78,9 +78,11 @@ aggregated_troll_data <-
     STEELHEAD_RELD = sum(STEELHEAD_RELD, na.rm = TRUE)
   )
 
+# Combine all tables
 aggregated_catch_data <- 
   bind_rows(aggregated_gill_net_data, aggregated_seine_data, aggregated_troll_data)
 
+# Sum up all catch statistics for any type of fishing
 aggregated_catch_data_by_area <-
   aggregated_catch_data |>
   group_by(CALENDAR_YEAR, MGMT_AREA) |>
@@ -139,6 +141,8 @@ write_parquet(pivoted_catch_data, "output/data/aggregated_pivoted_catch_data.par
 # Aggregate spawning population data
 
 spawning_population <- read_parquet('output/data/spawning_population.parquet')
+
+# Group first by the species, area, and year
 aggregated_spawning_population_by_area <- 
   spawning_population |>
   mutate(SALMON_POPULATION = coalesce(TOTAL_RETURN_TO_RIVER, NATURAL_ADULT_SPAWNERS)) |>
@@ -152,8 +156,10 @@ aggregated_spawning_population_by_area <-
     SALMON_POPULATION = sum(SALMON_POPULATION, na.rm = TRUE)
   )
 
+# Rename values to be capitalized
 aggregated_spawning_population_by_area$SPECIES = toupper(aggregated_spawning_population_by_area$SPECIES)
 
+# Group together only by species and year
 aggregated_spawning_population <- 
   aggregated_spawning_population_by_area |>
   group_by(SPECIES, YEAR) |>

### 2024-04-17 scripts/04.1_catch_model.R
model based on total catches instead

@@ -32,11 +32,25 @@ spawning_population_data <-
   filter(YEAR >= 2002) |>
   select(YEAR, SPECIES, SALMON_POPULATION)
 
+# Join the tables together by year and species
 model_data <- 
   left_join(catch_data, spawning_population_data, by = c("YEAR", "SPECIES")) |>
   mutate(SALMON_POPULATION = ifelse(is.na(SALMON_POPULATION), 0, SALMON_POPULATION)) |>
   filter(SPECIES != "STEELHEAD") # Not common
 
+# Gather the data to predict total salmon population and not just one species
+model_data <- 
+  model_data |>
+  group_by(YEAR) |>
+  summarize(
+    VESSEL_COUNT = max(VESSEL_COUNT), # Max since all are the same for a year
+    BOAT_DAYS = max(BOAT_DAYS),
+    SALMON_KPT = sum(SALMON_KPT),
+    SALMON_RLD = sum(SALMON_RLD),
+    SALMON_POPULATION = sum(SALMON_POPULATION)
+  )
+
+# The linear model based on the catch data
 catch_population_model <- lm(SALMON_POPULATION ~ VESSEL_COUNT + BOAT_DAYS + SALMON_KPT + SALMON_RLD, data = model_data)
 
 saveRDS(

### 2024-04-17 scripts/03_aggregate_data.R
update model to use standardized values and the percentage of catches coming from fishing method

@@ -20,7 +20,7 @@ troll_catches <- read_parquet("output/data/troll_catches.parquet")
 
 aggregated_gill_net_data <- 
   gill_net_catches |> 
-  group_by(CALENDAR_YEAR, MGMT_AREA) |>
+  group_by(CALENDAR_YEAR) |>
   summarize(
     VESSEL_COUNT = sum(VESSEL_COUNT, na.rm = TRUE),
     BOAT_DAYS = sum(BOAT_DAYS, na.rm = TRUE),
@@ -36,11 +36,12 @@ aggregated_gill_net_data <-
     CHINOOK_RELD = sum(CHINOOK_RELD, na.rm = TRUE),
     STEELHEAD_KEPT = sum(STEELHEAD_KEPT, na.rm = TRUE),
     STEELHEAD_RELD = sum(STEELHEAD_RELD, na.rm = TRUE)
-  )
+  )|>
+  mutate(KEPT_SUM = SOCKEYE_KEPT + COHO_KEPT + PINK_KEPT + CHUM_KEPT + CHINOOK_KEPT + STEELHEAD_KEPT)
 
 aggregated_seine_data <- 
   seine_catches |> 
-  group_by(CALENDAR_YEAR, MGMT_AREA) |>
+  group_by(CALENDAR_YEAR) |>
   summarize(
     VESSEL_COUNT = sum(VESSEL_COUNT, na.rm = TRUE),
     BOAT_DAYS = sum(BOAT_DAYS, na.rm = TRUE),
@@ -56,11 +57,12 @@ aggregated_seine_data <-
     CHINOOK_RELD = sum(CHINOOK_RELD, na.rm = TRUE),
     STEELHEAD_KEPT = sum(STEELHEAD_KEPT, na.rm = TRUE),
     STEELHEAD_RELD = sum(STEELHEAD_RELD, na.rm = TRUE)
-  )
+  )|>
+  mutate(KEPT_SUM = SOCKEYE_KEPT + COHO_KEPT + PINK_KEPT + CHUM_KEPT + CHINOOK_KEPT + STEELHEAD_KEPT)
 
 aggregated_troll_data <- 
   troll_catches |> 
-  group_by(CALENDAR_YEAR, MGMT_AREA) |>
+  group_by(CALENDAR_YEAR) |>
   summarize(
     VESSEL_COUNT = sum(VESSEL_COUNT, na.rm = TRUE),
[…]

### 2024-04-17 scripts/04.1_catch_model.R

@@ -25,6 +25,12 @@ catch_data <-
 
 spawning_population_data <- read_parquet("output/data/aggregated_spawning_population.parquet")
 
+catch_by_method <- read_parquet("output/data/catch_by_method.parquet")
+
+catch_by_method <-
+  catch_by_method |>
+  rename(YEAR = CALENDAR_YEAR)
+
 # Since for this situation, the catch data only starts at 2002, only consider 
 # spawning data for those years
 spawning_population_data <-
@@ -36,7 +42,8 @@ spawning_population_data <-
 model_data <- 
   left_join(catch_data, spawning_population_data, by = c("YEAR", "SPECIES")) |>
   mutate(SALMON_POPULATION = ifelse(is.na(SALMON_POPULATION), 0, SALMON_POPULATION)) |>
-  filter(SPECIES != "STEELHEAD") # Not common
+  filter(SPECIES != "STEELHEAD") |> # Not common 
+  left_join(catch_by_method, by = "YEAR")
 
 # Gather the data to predict total salmon population and not just one species
 model_data <- 
@@ -45,13 +52,16 @@ model_data <-
   summarize(
     VESSEL_COUNT = max(VESSEL_COUNT), # Max since all are the same for a year
     BOAT_DAYS = max(BOAT_DAYS),
-    SALMON_KPT = sum(SALMON_KPT),
-    SALMON_RLD = sum(SALMON_RLD),
-    SALMON_POPULATION = sum(SALMON_POPULATION)
+    SALMON_KPT = sum(SALMON_KPT) / 100000, # Track catches in 100,000s
+    SALMON_RLD = sum(SALMON_RLD) / 100000,
+    SALMON_POPULATION = sum(SALMON_POPULATION) / 100000,
+    PERCENTAGE_gill_net = max(PERCENTAGE_gill_net) * 100, # Convert decimals to percentages
+    PERCENTAGE_seine = max(PERCENTAGE_seine) * 100,
+    PERCENTAGE_troll = max(PERCENTAGE_troll) * 100
   )
 
 # The linear model based on the catch data
-catch_population_model <- lm(SALMON_POPULATION ~ VESSEL_COUNT + BOAT_DAYS + SALMON_KPT + SALMON_RLD, data = model_data)
[…]

### 2024-04-18 scripts/01_download_data.R
use readxl to properly handle date values

@@ -9,7 +9,8 @@
 #### Workplace setup ####
 
 library(tidyverse)
-library(openxlsx)
+library(readxl)
+library(lubridate)
 
 #### Download and save Salmon data ####
 
@@ -26,7 +27,7 @@ unzip(zip_file, exdir="input/data/")
 url <- "https://api-proxy.edh.azure.cloud.dfo-mpo.gc.ca/catalogue/records/c48669a3-045b-400d-b730-48aafe8c5ee6/attachments/All%20Areas%20NuSEDS_20240110.xlsx"
 download.file(url, "input/data/nuseds.xlsx", mode = "wb")
 
-nuseds_data_xlsx <- read.xlsx("input/data/nuseds.xlsx")
+nuseds_data_xlsx <- read_excel("input/data/nuseds.xlsx")
 
 # Filter only relevant columns, otherwise the file becomes too large to save to github
 nuseds_data_xlsx <- 
@@ -35,5 +36,7 @@ nuseds_data_xlsx <-
          -NATURAL_ADULT_FEMALES, -NATURAL_ADULT_MALES, -EFFECTIVE_FEMALES, -WEIGHTED_PCT_SPAWN,
          -WATERSHED_CDE, -POPULATION, -ACCURACY, -PRECISION, -INDEX_YN, -RELIABILITY, 
          -ESTIMATE_STAGE, -ESTIMATE_CLASSIFICATION, -NO_INSPECTIONS_USED, -ESTIMATE_METHOD, -CREATED_DTT, -UPDATED_DTT)
+
+nuseds_data_xlsx$START_DTT <- ymd(nuseds_data_xlsx$START_DTT)
 
 write_csv(nuseds_data_xlsx, "input/data/nuseds.csv")

### 2024-04-18 scripts/02.2_clean_spawn_data.R
separate aggregations

@@ -38,10 +38,12 @@ spawning_population <-
 # Broodstock (amount taken out for the use of human influenced spawning)
 broodstock <-
   nuseds_data |> 
-  select(AREA, WATERBODY, LOCAL_NAME, YEAR, SPECIES,
+  select(AREA, WATERBODY, YEAR, SPECIES,
          ADULT_BROODSTOCK_REMOVALS, JACK_BROODSTOCK_REMOVALS, TOTAL_BROODSTOCK_REMOVALS, OTHER_REMOVALS, TOTAL_RETURN_TO_RIVER) |>
-  filter(!is.na(ADULT_BROODSTOCK_REMOVALS) & !is.na(TOTAL_BROODSTOCK_REMOVALS) & !is.na(TOTAL_RETURN_TO_RIVER))
-
+  filter(!is.na(ADULT_BROODSTOCK_REMOVALS) & !is.na(TOTAL_BROODSTOCK_REMOVALS) & !is.na(TOTAL_RETURN_TO_RIVER)) |>
+  mutate(JACK_BROODSTOCK_REMOVALS = ifelse(is.na(JACK_BROODSTOCK_REMOVALS), 0, JACK_BROODSTOCK_REMOVALS)) |>
+  mutate(OTHER_REMOVALS = ifelse(is.na(OTHER_REMOVALS), 0, OTHER_REMOVALS))
+  
 # Spawn timing features (arrival, peak)
 spawn_timings <-
   nuseds_data |> 

### 2024-04-18 scripts/03.2_aggregated_spawn_data.R

@@ -1,0 +1,55 @@
+#### Preamble ####
+# Purpose: Group together statistics based on species
+# Author: Timothius Prajogi
+# Date: 18 April 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01_download_data, 02.x_clean_data
+
+#### Workplace setup ####
+
+library(tidyverse)
+library(arrow)
+
+#### Aggregate By Species ####
+
+# Load in data
+spawning_population <- read_parquet('output/data/spawning_population.parquet')
+broodstock <- read_parquet("output/data/broodstock.parquet")
+
+# Aggregate spawning population data
+
+# Group first by the species, area, and year
+aggregated_spawning_population_by_area <- 
+  spawning_population |>
+  mutate(SALMON_POPULATION = coalesce(TOTAL_RETURN_TO_RIVER, NATURAL_ADULT_SPAWNERS)) |>
+  filter(SPECIES != "Atlantic" & SPECIES != "Kokanee") |>
+  group_by(SPECIES, AREA, YEAR) |>
+  summarize(
+    NATURAL_ADULT_SPAWNERS = sum(NATURAL_ADULT_SPAWNERS, na.rm = TRUE),
+    NATURAL_JACK_SPAWNERS = sum(NATURAL_JACK_SPAWNERS, na.rm = TRUE),
+    NATURAL_SPAWNERS_TOTAL = sum(NATURAL_SPAWNERS_TOTAL, na.rm = TRUE),
+    TOTAL_RETURN_TO_RIVER = sum(TOTAL_RETURN_TO_RIVER, na.rm = TRUE),
+    SALMON_POPULATION = sum(SALMON_POPULATION, na.rm = TRUE)
+  )
+
+# Rename values to be capitalized
+aggregated_spawning_population_by_area$SPECIES = toupper(aggregated_spawning_population_by_area$SPECIES)
+
+# Group together only by species and year
+aggregated_spawning_population <- 
+  aggregated_spawning_population_by_area |>
+  group_by(SPECIES, YEAR) |>
+  summarize(
+    NATURAL_ADULT_SPAWNERS = sum(NATURAL_ADULT_SPAWNERS, na.rm = TRUE),
+    NATURAL_JACK_SPAWNERS = sum(NATURAL_JACK_SPAWNERS, na.rm = TRUE),
[…]

### 2024-04-18 scripts/02.2_clean_spawn_data.R
group and change the way na values are handled for broodstock

@@ -39,10 +39,19 @@ spawning_population <-
 broodstock <-
   nuseds_data |> 
   select(AREA, WATERBODY, YEAR, SPECIES,
+         NATURAL_ADULT_SPAWNERS, NATURAL_SPAWNERS_TOTAL,
          ADULT_BROODSTOCK_REMOVALS, JACK_BROODSTOCK_REMOVALS, TOTAL_BROODSTOCK_REMOVALS, OTHER_REMOVALS, TOTAL_RETURN_TO_RIVER) |>
-  filter(!is.na(ADULT_BROODSTOCK_REMOVALS) & !is.na(TOTAL_BROODSTOCK_REMOVALS) & !is.na(TOTAL_RETURN_TO_RIVER)) |>
-  mutate(JACK_BROODSTOCK_REMOVALS = ifelse(is.na(JACK_BROODSTOCK_REMOVALS), 0, JACK_BROODSTOCK_REMOVALS)) |>
-  mutate(OTHER_REMOVALS = ifelse(is.na(OTHER_REMOVALS), 0, OTHER_REMOVALS))
+  filter(!is.na(NATURAL_ADULT_SPAWNERS) | !is.na(NATURAL_SPAWNERS_TOTAL) | !is.na(TOTAL_RETURN_TO_RIVER)) |>
+  mutate(
+    NATURAL_ADULT_SPAWNERS = ifelse(is.na(NATURAL_ADULT_SPAWNERS), 0, NATURAL_ADULT_SPAWNERS),
+    NATURAL_SPAWNERS_TOTAL = ifelse(is.na(NATURAL_SPAWNERS_TOTAL), 0, NATURAL_SPAWNERS_TOTAL),
+    TOTAL_RETURN_TO_RIVER = ifelse(is.na(TOTAL_RETURN_TO_RIVER), 0, TOTAL_RETURN_TO_RIVER),
+    ADULT_BROODSTOCK_REMOVALS = ifelse(is.na(ADULT_BROODSTOCK_REMOVALS), 0, ADULT_BROODSTOCK_REMOVALS),
+    JACK_BROODSTOCK_REMOVALS = ifelse(is.na(JACK_BROODSTOCK_REMOVALS), 0, JACK_BROODSTOCK_REMOVALS),
+    TOTAL_BROODSTOCK_REMOVALS = ifelse(is.na(TOTAL_BROODSTOCK_REMOVALS), 0, TOTAL_BROODSTOCK_REMOVALS),
[…]

### 2024-04-18 scripts/03.2_aggregated_spawn_data.R

@@ -52,4 +52,17 @@ write_parquet(aggregated_spawning_population, "output/data/aggregated_spawning_p
 
 # Aggregate broodstock data
 
+aggregated_broodstock <-
+  broodstock |>
+  group_by(YEAR) |>
+  summarize(
+    ADULT_BROODSTOCK_REMOVALS = sum(ADULT_BROODSTOCK_REMOVALS, na.rm = TRUE),
+    JACK_BROODSTOCK_REMOVALS = sum(JACK_BROODSTOCK_REMOVALS, na.rm = TRUE),
+    TOTAL_BROODSTOCK_REMOVALS = sum(TOTAL_BROODSTOCK_REMOVALS, na.rm = TRUE),
+    OTHER_REMOVALS = sum(OTHER_REMOVALS, na.rm = TRUE),
+    TOTAL_RETURN_TO_RIVER = sum(TOTAL_RETURN_TO_RIVER, na.rm = TRUE)
+  ) |>
+  filter(YEAR >= 1979) |>
+  arrange(YEAR)
 
+write_parquet(aggregated_broodstock, "output/data/aggregated_broodstock.parquet")

### 2024-04-18 scripts/04.1_catch_model.R
Model the yearly changes of population

@@ -8,7 +8,6 @@
 
 #### Workplace setup ####
 
-library(rstanarm)
 library(tidyverse)
 library(modelsummary)
 

### 2024-04-18 scripts/04.2_yearly_change_model.R

@@ -1,0 +1,50 @@
+#### Preamble ####
+# Purpose: Create LM predicting Salmon spawn using previous years data
+# Author: Timothius Prajogi
+# Date: 18 April 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01_download_data, 02.x_clean_data, 03_aggregate_data
+
+#### Workplace setup ####
+
+library(rstanarm)
+library(tidyverse)
+library(modelsummary)
+
+set.seed(302)
+
+#### Create Model ####
+
+# Load data
+broodstock <- read_parquet("output/data/aggregated_broodstock.parquet")
+
+broodstock_next_year <- 
+  broodstock |>
+  select(YEAR, TOTAL_RETURN_TO_RIVER) |>
+  mutate(YEAR = YEAR - 1) |>
+  rename(NEXT_YEAR_RETURN = TOTAL_RETURN_TO_RIVER)
+
+# Since for the first and last year, we have nothing to predict, we remove those
+broodstock <- 
+  broodstock |>
+  filter(YEAR != max(YEAR))
+
+broodstock_next_year <-
+  broodstock_next_year |>
+  filter(YEAR != min(YEAR))
+
+# Join the tables (essentially using last years spawning data to predict the next years spawning data)
+model_data <- 
+  left_join(broodstock, broodstock_next_year, by = "YEAR") |>
+  mutate(
+    TOTAL_RETURN_TO_RIVER = TOTAL_RETURN_TO_RIVER / 1000,
+    NEXT_YEAR_RETURN = NEXT_YEAR_RETURN / 1000
+  )
+
+yearly_change_model <- lm(NEXT_YEAR_RETURN ~ ADULT_BROODSTOCK_REMOVALS + JACK_BROODSTOCK_REMOVALS + TOTAL_BROODSTOCK_REMOVALS + OTHER_REMOVALS + TOTAL_RETURN_TO_RIVER, data = model_data)
+
+saveRDS(
+  yearly_change_model,
+  file = "output/models/yearly_change_model.rds"
+)

### 2024-04-18 scripts/04.1_catch_model.R
add the model library

@@ -8,6 +8,7 @@
 
 #### Workplace setup ####
 
+library(rstanarm)
 library(tidyverse)
 library(modelsummary)
 

### 2024-04-18 output/paper.qmd
Update intro and abstract to match changes

@@
 date-format: long
-abstract: "Using the NuSEDS dataset provided by the Government of Canada and statistics published by Fisheries and Oceans Canada regarding the number of commercial fishing licences issued by Pacific Region, we investigate the effects of commercial fishing on the reproductive capacity of Canadian Salmon.
-In tandem, we investigate how much of the change in spawn rates can be attributed to the health of the ecosystem, climate, and the fish that live in it.
+abstract: "Using the NuSEDS dataset and Catch Estimates for the Pacific Region Commerical Salmon Fishery dataset provided by the Government of Canada, we investigate the effects of commercial fishing on the reproductive capacity of Canadian Salmon.
+Along with that, we investigate how much of the change in spawn rates can be attributed to the health of the ecosystem, climate, and the fish that live in it.
 This paper discusses the impact of direct human intervention in the form of commercial fishing, aiming to highlight the importance in proper management and control in the fishing industry in alignement with changes in the environment by predicting spawning trends of future generations."
@@
 This paper will focus on commercial fishing trends, ocean climate trends, and spawning rates over the last decade, using datasets and fisheries statistics provided through the Government of Canada's open data portal.
-In particular, the NuSEDS (New Salmon Escapement Database System) provides key insights on spawner records and abundance estimates for freshwater streams and tributaries, with statistics provided by Fisheries and Oceans Canada providing detail on the amount of licenses issued and fish landed by species.
-Through this data we discover the potential drivers of salmon population decline, including but not limited to habitat degradation, climate change impacts, and human-induced factors such as over fishing and habitat loss.
+In particular, the NuSEDS (New Salmon Escapement Database System) provides key insights on spawner records and abundance estimates for freshwater streams and tributaries, with statistics provided by Fisheries and Oceans Canada providing detail on the amount of fish landed or released by species.
+Through this data we discover the potential human-induced drivers of salmon population change including factors such as over fishing and artificial breeding.
 In identifying the critical areas for repair and reform, we aim to explore potential mitigation strategies and management plans with the goal of restoring salmon populations and promoting sustainable fishing practices, protecting this key resource.
@@
 
-This paper will introduce the data and models we used to analyze these themes and trends, presenting our findings and their implications for future salmon generations.
+Through information gathered from these datasets, we were able to verify correlations between the frequency of fishing and the salmon population, as well as the previous years salmon population and artificial breeding practices affect on future generations.
+However, although these are significant factors, they do not account for all the change that is seen throughout the years.
+Salmon population's fluctuation cannot be fully explained by these human factors alone, which means there must be some other factors that are not described in this data that should be investigated.
+
+This paper will introduce the data and models we used to analyze these themes and trends, explaining why features and datasets were chosen and how they are relevant to the research topic.
+Then the paper will present our findings and their implications for future salmon generations observed in the models that were created as well as in the data.
+We will conclude by discussing the weaknesses and future steps that could be made to further improve on this research.
 

### 2024-04-18 scripts/02.2_clean_spawn_data.R
clean incorrect column

@@ -33,7 +33,11 @@ spawning_population <-
   nuseds_data |>
   select(AREA, WATERBODY, LOCAL_NAME, YEAR, SPECIES,
          NATURAL_ADULT_SPAWNERS, NATURAL_JACK_SPAWNERS, NATURAL_SPAWNERS_TOTAL, TOTAL_RETURN_TO_RIVER) |>
-  filter(!is.na(NATURAL_ADULT_SPAWNERS))
+  filter(!is.na(NATURAL_ADULT_SPAWNERS)) |>
+  mutate(
+    NATURAL_ADULT_SPAWNERS = ifelse(is.na(NATURAL_ADULT_SPAWNERS), 0, NATURAL_ADULT_SPAWNERS),
+    NATURAL_JACK_SPAWNERS = ifelse(is.na(NATURAL_JACK_SPAWNERS), 0, NATURAL_JACK_SPAWNERS)
+  )
   
 # Broodstock (amount taken out for the use of human influenced spawning)
 broodstock <-
@@ -51,7 +55,7 @@ broodstock <-
     TOTAL_BROODSTOCK_REMOVALS = ifelse(is.na(TOTAL_BROODSTOCK_REMOVALS), 0, TOTAL_BROODSTOCK_REMOVALS),
     OTHER_REMOVALS = ifelse(is.na(OTHER_REMOVALS), 0, OTHER_REMOVALS),
   ) |>
-  mutate(TOTAL_RETURN_TO_RIVER, max(NATURAL_ADULT_SPAWNERS, NATURAL_SPAWNERS_TOTAL, TOTAL_RETURN_TO_RIVER))
+  mutate(TOTAL_RETURN_TO_RIVER = pmax(NATURAL_ADULT_SPAWNERS, NATURAL_SPAWNERS_TOTAL, TOTAL_RETURN_TO_RIVER))
   
 # Spawn timing features (arrival, peak)
 spawn_timings <-

### 2024-04-18 scripts/04.1_catch_model.R
update models to use stan_glm

@@ -52,16 +52,25 @@ model_data <-
   summarize(
     VESSEL_COUNT = max(VESSEL_COUNT), # Max since all are the same for a year
     BOAT_DAYS = max(BOAT_DAYS),
-    SALMON_KPT = sum(SALMON_KPT) / 100000, # Track catches in 100,000s
-    SALMON_RLD = sum(SALMON_RLD) / 100000,
-    SALMON_POPULATION = sum(SALMON_POPULATION) / 100000,
+    SALMON_KPT = sum(SALMON_KPT), # Track catches in 100,000s
+    SALMON_RLD = sum(SALMON_RLD),
+    SALMON_POPULATION = sum(SALMON_POPULATION),
     PERCENTAGE_gill_net = max(PERCENTAGE_gill_net) * 100, # Convert decimals to percentages
     PERCENTAGE_seine = max(PERCENTAGE_seine) * 100,
     PERCENTAGE_troll = max(PERCENTAGE_troll) * 100
   )
 
 # The linear model based on the catch data
-catch_population_model <- lm(SALMON_POPULATION ~ VESSEL_COUNT + BOAT_DAYS + SALMON_KPT + SALMON_RLD + PERCENTAGE_gill_net + PERCENTAGE_seine + PERCENTAGE_troll, data = model_data)
+catch_population_model <- 
+  stan_glm(
+    formula = SALMON_POPULATION ~ VESSEL_COUNT + BOAT_DAYS + SALMON_KPT + SALMON_RLD + PERCENTAGE_gill_net + PERCENTAGE_seine + PERCENTAGE_troll, 
+    data = model_data,
+    family = gaussian(),
+    prior = normal(location = 8, scale = 2.5, autoscale = TRUE),
+    prior_intercept = normal(0, 2.5, autoscale = TRUE),
+    prior_aux = exponential(rate = 1, autoscale = TRUE),
+    seed = 302
+  )
 
 saveRDS(
   catch_population_model,

### 2024-04-18 scripts/04.2_yearly_change_model.R

@@ -42,7 +42,16 @@ model_data <-
     NEXT_YEAR_RETURN = NEXT_YEAR_RETURN / 1000
   )
 
-yearly_change_model <- lm(NEXT_YEAR_RETURN ~ ADULT_BROODSTOCK_REMOVALS + JACK_BROODSTOCK_REMOVALS + TOTAL_BROODSTOCK_REMOVALS + OTHER_REMOVALS + TOTAL_RETURN_TO_RIVER, data = model_data)
+yearly_change_model <- 
+  stan_glm(
+    formula = NEXT_YEAR_RETURN ~ ADULT_BROODSTOCK_REMOVALS + JACK_BROODSTOCK_REMOVALS + TOTAL_BROODSTOCK_REMOVALS + OTHER_REMOVALS + TOTAL_RETURN_TO_RIVER,
+    data = model_data,
+    family = gaussian(),
+    prior = normal(location = 8, scale = 2.5, autoscale = TRUE),
+    prior_intercept = normal(0, 2.5, autoscale = TRUE),
+    prior_aux = exponential(rate = 1, autoscale = TRUE),
+    seed = 302
+  )
 
 saveRDS(
   yearly_change_model,

### 2024-04-18 scripts/04.1_catch_model.R
Update models to use stan_glm

@@ -54,7 +54,7 @@ model_data <-
     BOAT_DAYS = max(BOAT_DAYS),
     SALMON_KPT = sum(SALMON_KPT), # Track catches in 100,000s
     SALMON_RLD = sum(SALMON_RLD),
-    SALMON_POPULATION = sum(SALMON_POPULATION),
+    SALMON_POPULATION = sum(SALMON_POPULATION) / 1000,
     PERCENTAGE_gill_net = max(PERCENTAGE_gill_net) * 100, # Convert decimals to percentages
     PERCENTAGE_seine = max(PERCENTAGE_seine) * 100,
     PERCENTAGE_troll = max(PERCENTAGE_troll) * 100
@@ -66,7 +66,7 @@ catch_population_model <-
     formula = SALMON_POPULATION ~ VESSEL_COUNT + BOAT_DAYS + SALMON_KPT + SALMON_RLD + PERCENTAGE_gill_net + PERCENTAGE_seine + PERCENTAGE_troll, 
     data = model_data,
     family = gaussian(),
-    prior = normal(location = 8, scale = 2.5, autoscale = TRUE),
+    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_intercept = normal(0, 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
     seed = 302

### 2024-04-18 scripts/04.2_yearly_change_model.R

@@ -47,7 +47,7 @@ yearly_change_model <-
     formula = NEXT_YEAR_RETURN ~ ADULT_BROODSTOCK_REMOVALS + JACK_BROODSTOCK_REMOVALS + TOTAL_BROODSTOCK_REMOVALS + OTHER_REMOVALS + TOTAL_RETURN_TO_RIVER,
     data = model_data,
     family = gaussian(),
-    prior = normal(location = 8, scale = 2.5, autoscale = TRUE),
+    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_intercept = normal(0, 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
     seed = 302

### 2024-04-18 output/paper.qmd
Complete data section nuseds

@@
 
+```{r}
+#| include: false
+#| warning: false
+#| message: false
+#| echo: false
+
+#### Workspace setup ####
+library(knitr)
+library(tidyverse)
+library(modelsummary)
+library(rstanarm)
+library(arrow)
+```
+
 # Introduction
@@
 This paper will focus on commercial fishing trends, ocean climate trends, and spawning rates over the last decade, using datasets and fisheries statistics provided through the Government of Canada's open data portal.
-In particular, the NuSEDS (New Salmon Escapement Database System) provides key insights on spawner records and abundance estimates for freshwater streams and tributaries, with statistics provided by Fisheries and Oceans Canada providing detail on the amount of fish landed or released by species.
+In particular, the NuSEDS (New Salmon Escapement Database System) provides key insights on spawn records and abundance estimates for freshwater streams and tributaries, with statistics provided by Fisheries and Oceans Canada providing detail on the amount of fish landed or released through commercial fishing methods by species.
 Through this data we discover the potential human-induced drivers of salmon population change including factors such as over fishing and artificial breeding.
@@
 This paper will introduce the data and models we used to analyze these themes and trends, explaining why features and datasets were chosen and how they are relevant to the research topic.
+The models we create will use commercial fisheries' catch estimates and past years artificial and natural spawn success respectively as the estimands to predict salmon populations.
 Then the paper will present our findings and their implications for future salmon generations observed in the models that were created as well as in the data.
@@
 
+The data used in this paper was provided by the Government of Canada through their Open Government Portal, published by Fisheries and Oceans Canada.
+
+For information regarding the populations of Canadian Salmon in the Pacific region, the New Salmon Escapement Database System was used to retrieve the quality and health of the Salmon population.
+This data is not only used to measure the populations but other health and success measures that certain area salmon have.
+These observations are made during the spawning season for salmon (from September - November), recording the amount of salmon that have come to spawn in a certain waterbody.
+As it is not possible for those wildlife officials to individually count each salmon that enters a waterway at any time during the day, an estimated count was made based on the abundance of salmon currently present during the observation.
+
+The database provides salmon spawning observations starting from 1920's, but population numbers were not recorded / provided until 1979.
+As such, rows not providing population numbers were not considered.
+In addition, for each model, specific features were necessary to create the model, and the original dataset was split into two models, each only containing the entries that include relevant information.
+
+```{r}
+#| include: false
+#| warning: false
+#| message: false
+#| echo: false
+
+# Load in data
+
+# Catch data
+gill_net_catches <- read_parquet("data/gill_net_catches.parquet")
+seine_catches <- read_parquet("data/seine_catches.parquet")
+troll_catches <- read_parquet("data/troll_catches.parquet")
+
+# Spawn data
+spawning_population <- read_parquet("data/spawning_population.parquet")
+broodstock <- read_parquet("data/broodstock.parquet")
+
+```
+
+```{r, table, width="100%"}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: tbl-spawnpopulation
+#| tbl-cap: "NuSEDS Data on Spawning Population"
+
+spawning_population |>
+    select(-LOCAL_NAME, -NATURAL_SPAWNERS_TOTAL) |>
+    head() |>
+    kable(
+      col.names = c("Area", "Name of Waterbody", "Year", "Species", "Natural Adult Spawners", "Natural Jack Spawners", "Total Return To River"),
+      booktabs = TRUE
+    )
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: tbl-broodstock
+#| tbl-cap: "NuSEDS Data on Broodstock and Population"
+
+broodstock |>
+  select(-NATURAL_ADULT_SPAWNERS, -NATURAL_SPAWNERS_TOTAL, -TOTAL_BROODSTOCK_REMOVALS) |>
+  filter(ADULT_BROODSTOCK_REMOVALS > 0) |> # For the sack of the visual
+  head() |>
+  kable(
+    col.names = c("Area", "Name of Waterbody", "Year", "Species", "Adult Broodstock Removals", "Jack Broodstock Removals", "Other Removals", "Total Return To River"),
+    booktabs = TRUE
+  )
+```
+
+The most important part of this dataset that we consider is the spawning salmon population observed at various waterways as shown in @tbl-spawnpopulation.
+We then also consider the broodstock removal statistics in @tbl-broodstock, which is the amount of salmon that were removed for the purposes of artificial breeding.
+These salmon's progeny are later released back into the wild.
+
+Variables for both @tbl-spawnpopulation and @tbl-broodstock:
+
+Waterbody + Area + Year:
+
+  -   This defines the waterbody and the general management area that this observation came from, and what year that observation was made.
+I have provided the map that shows the actual geographical area that an area code signifies.
+
+Species:
+
+  -   This defines the species of salmon that the observation was made for.
+Observations were made for Sockeye, Chum, Pink, chinook, Coho, Steelhead, Atlantic, and Kokanee salmon, but only observations for the more common salmon species were made.
+(Sockeye, Chum, Pink, Chinook, and Coho) These are the salmon that make up the vast majority of the sport and the commercial market, which is the focus of this paper, and why the other salmon observations were not considered.
+
+Total Return to River:
+
+  -   This defines the total amount of salmon of that species that was observed in that waterbody, being number of natural spawners + number of non-natural spawners (fish released / cultivated by the government to support salmon population).
+For @tbl-spawnpopulation, an N/A value indicates that only natural spawners were observed.
+
+Variables for @tbl-spawnpopulation:
+
+Natural Adult Spawners:
+
+  -   These are the amount of salmon observed that have reached maturity.
+
+Natural Jack Spawners:
+
+  -   These are fish that have matured at an early age, and are much smaller than the adult spawners. When the value is N/A this indicates that no jacks were not observed.
+
+Variables for @tbl-broodstock:
+
+Adult Broodstock Removals:
+
+  -   Similarly to the spawning population data, these are the amount of mature salmon that were removed from the waterbody for the purposes of breeding.
+
+Jack Broodstock Removals:
+
+  -   Similarly to the spawning population data, these are the amount of jack salmon that were removed from the waterbody for the purposes of breeding.
+
+Other Removals:
+
+  -   These are the amount of salmon that were removed from their natural environment by humans for other reasons (not broodstock) which were not defined by the dataset.
+
+For information regarding
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: tbl-catchdata_gill
+#| tbl-cap: "Commercial Salmon Fishing Catch Estimates from 2005 - 2022 from Gill Net Fisheries"
+
+gill_net_catches |>
+  select(-FISHERY, -LICENCE_AREA, -NOTES) |>
+  head() |>
+  kable(
+    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released", "Coho Kept", "Coho Released", "Pink Kept", "Pink Released", "Chum Kept", "Chum Released", "Chinook Kept", "Chinook Released", "Steelhead Kept", "Steelhead Released"),
+    booktabs = TRUE
+  )
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: tbl-catchdata_seine
+#| tbl-cap: "Commercial Salmon Fishing Catch Estimates from 2005 - 2022 from Seine Net Fisheries"
+
+seine_catches |>
+  select(-FISHERY, -LICENCE_AREA, -NOTES) |>
+  head() |>
+  kable(
+    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released", "Coho Kept", "Coho Released", "Pink Kept", "Pink Released", "Chum Kept", "Chum Released", "Chinook Kept", "Chinook Released", "Steelhead Kept", "Steelhead Released"),
+    booktabs = TRUE
+  )
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: tbl-catchdata_troll
+#| tbl-cap: "Commercial Salmon Fishing Catch Estimates from 2005 - 2022 from Trolling Fisheries"
+
+troll_catches |>
+  select(-FISHERY, -LICENCE_AREA, -NOTES) |>
+  head() |>
+  kable(
+    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released", "Coho Kept", "Coho Released", "Pink Kept", "Pink Released", "Chum Kept", "Chum Released", "Chinook Kept", "Chinook Released", "Steelhead Kept", "Steelhead Released"),
+    booktabs = TRUE
+  )
+```
+
 # Model
@@
 
+```{r}
+# TODO potentially plot the spawning peak dates vs year to show if seasons affect it
+```
+
 # Discussion
+
+```{r}
+# TODO talk about the fact that region caught vs population count region were not considered,
+# to be more accuracte, we would monitor changes between each watershed
+
+# Talk about the large amount of na values, population may not be accurately represented in some years
+```
 

### 2024-04-19 output/paper.qmd
Finish data section (-justification and relationships)

@@
 
-For information regarding the populations of Canadian Salmon in the Pacific region, the New Salmon Escapement Database System was used to retrieve the quality and health of the Salmon population.
+## Spawn
+
+For information regarding the populations of Canadian Salmon in the Pacific region, the *New Salmon Escapement Database System (NuSEDS)* was used to retrieve the quality and health of the Salmon population.
 This data is not only used to measure the populations but other health and success measures that certain area salmon have.
@@
 
-For information regarding
+## Catch
+
+For information regarding the salmon caught in commercial fisheries, the *Pacific Region Commercial Salmon Fishery In-season Catch Estimates* was used.
+This data comes from the Fishery Operations System (FOS) which serves as the primary database for Fisheries and Oceans Canada's (DFO) Pacific Region Commercial Salmon Logbook program, established in 1998 and mandated fleet-wide participation by 2001.
+This program compels commercial salmon fishers to log daily catch and effort data in harvest logbooks, subsequently reporting it to DFO via service providers within specified deadlines.
+However, certain personal information provided by fishers is protected and inaccessible.
+Fishery Managers utilize this reported data, alongside additional sources like overflights, to compute in-season catch estimates.
+Its important to note that this dataset omits data from test fishing, recreational fishing, or First Nations harvest.
 
@@
 gill_net_catches |>
-  select(-FISHERY, -LICENCE_AREA, -NOTES) |>
+  select(CALENDAR_YEAR, MGMT_AREA, VESSEL_COUNT, BOAT_DAYS, SOCKEYE_KEPT, SOCKEYE_RELD) |>
   head() |>
   kable(
-    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released", "Coho Kept", "Coho Released", "Pink Kept", "Pink Released", "Chum Kept", "Chum Released", "Chinook Kept", "Chinook Released", "Steelhead Kept", "Steelhead Released"),
+    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released"),
     booktabs = TRUE
@@
 seine_catches |>
-  select(-FISHERY, -LICENCE_AREA, -NOTES) |>
+  select(CALENDAR_YEAR, MGMT_AREA, VESSEL_COUNT, BOAT_DAYS, SOCKEYE_KEPT, SOCKEYE_RELD) |>
   head() |>
   kable(
-    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released", "Coho Kept", "Coho Released", "Pink Kept", "Pink Released", "Chum Kept", "Chum Released", "Chinook Kept", "Chinook Released", "Steelhead Kept", "Steelhead Released"),
+    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released"),
     booktabs = TRUE
@@
 troll_catches |>
-  select(-FISHERY, -LICENCE_AREA, -NOTES) |>
+  select(CALENDAR_YEAR, MGMT_AREA, VESSEL_COUNT, BOAT_DAYS, SOCKEYE_KEPT, SOCKEYE_RELD) |>
   head() |>
   kable(
-    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released", "Coho Kept", "Coho Released", "Pink Kept", "Pink Released", "Chum Kept", "Chum Released", "Chinook Kept", "Chinook Released", "Steelhead Kept", "Steelhead Released"),
+    col.names = c("Year", "Management Area", "Vessel Count", "Boat Days", "Sockeye Kept", "Sockeye Released"),
     booktabs = TRUE
@@
 
+This dataset is provided in three parts, separated by the type of commercial fishing that was being tracked. The three types being: Gill nets, Seine nets, and Trolling.
+
+**Gill Nets**
+
+  -   These are vertical panels of netting suspended in the water by a float line and a weighted line.
+Fish swimming into the net become entangled by their gills as they attempt to pass through the mesh.
+Gill nets are often set in a stationary position and left for a period of time before retrieval.
+
+**Seine Nets**
+
+  -   Seine nets are large, horizontal nets deployed from the side of a vessel.
+They typically have floats along the top edge and weights along the bottom to create a wall of netting that can be drawn in a circular motion to encircle fish.
+Once the fish are contained within the net, it is hauled in and the catch is retrieved.
+
+**Trolling**
+
+  -   Trolling involves trailing baited fishing lines behind a moving vessel.
+The lines are attached to fishing rods or reels and are drawn through the water at varying depths and speeds.
+This method allows fishermen to target specific species of fish by adjusting the depth and speed of the trolling lines.
+
+Understanding the type of fishing method used is crucial because each method has different impacts on salmon populations. For example:
+
+- Gill nets are known for their high selectivity, catching fish primarily by their gills. However, they can also unintentionally catch non-target species and may lead to bycatch issues.
+
+- Seine nets are effective for catching large quantities of fish in a single haul, but they may have a significant impact on fish habitat and can result in the incidental capture of non-target species.
+
+- Trolling is often considered a more sustainable fishing method compared to nets because it allows for the release of non-target species and reduces habitat damage.
+However, it can still have localized impacts on salmon populations if not properly managed.
+
+By categorizing the dataset based on the type of commercial fishing method, we can better understand the specific impacts of each method on salmon populations and inform fisheries management decisions accordingly.
+
+For each of @tbl-catchdata_gill, @tbl-catchdata_seine, @tbl-catchdata_troll, we consider the following variables:
+
+Year + Management Area:
+
+- Similarly to @tbl-spawnpopulation, this signifies the year and area the catch estimates were made for.
+However in this case the management area is different to the area mentioned prior.
+It also refers to a region, however this time it refers to the area that a fish was caught in.
+
+- Regarding this field, it was originally intended to be used to match up certain fishing regions to certain spawning regions.
+However, since it is not guaranteed or even expected that salmon stay close to their spawning grounds during the open fishing season, these area fields were not used in this matter.
+They do still have value that is not used in this report, as will be mentioned in the [discussion](#discussion) section.
+
+Vessel Count:
+
+- This measures the amount of vessels fishing in a certain management area in the year.
+
+Boat Days:
+
+- This measures the total amount of days that each vessel was fishing in that management area. For instance if there were two vessels fishing for the same two days each, that would be 4 boat days.
+
+Salmon Species Kept:
+
+- These are the amount of salmons of that particular species indicated in the column name were kept from that management area in that year.
+
+- Although @tbl-catchdata_gill, @tbl-catchdata_seine, and @tbl-catchdata_troll only show sockeye kept, the catch estimates for all other salmon species are also present, they were omitted here due for simplicity.
+
+Salmon Species Released:
+
+- These are the amount of salmons of that particular species indicated in the column name were caught but not kept and released back into the water from that management area in that year.
+
+- Although @tbl-catchdata_gill, @tbl-catchdata_seine, and @tbl-catchdata_troll only show sockeye kept, the catch estimates for all other salmon species are also present, they were omitted here due for simplicity.
+
+```{r}
+# TODO show all the relationships (broodstock vs population, total catches vs population, type vs population)
+```
+
 # Model
+
+## Catch Model
+
+## Yearly Model
 
@@
 
-```{r}
-# TODO potentially plot the spawning peak dates vs year to show if seasons affect it
-```
-
-# Discussion
+# Discussion {#discussion}
 
@@
 
-# Appendix
-
 # References

### 2024-04-19 output/paper.qmd
Add citations

@@
 The data used in this paper was provided by the Government of Canada through their Open Government Portal, published by Fisheries and Oceans Canada.
+This portal provided other datasets relating to salmon health / population, however, the datasets selected are consistently updated and provide relevant as well as historical data that is useful when researching yearly trends which salmon populations are due to the fact that future populations are largely impacted by the success of each year's spawn.
 
@@
 
+This data was downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized:
+
+-   `ggplot2` [@citeggplot]
+-   `dplyr` [@citedplyr]
+-   `readr` [@citereadr]
+-   `tibble` [@citetibble]
+
+For additional assistance with data storage, the `arrow` [@citearrow] was used and for report generation, the `knitr` [@citeknitr] package was used.
+
 # Model
+
+To create the linear model we made use of the `rstanarm` [@citerstanarm] package.
 

### 2024-04-19 output/paper.qmd
add remaining citations

@@
 
-The declining trend in Canadian salmon runs, as observed by many avid anglers and supported by empirical data (INSERT CITATION), serves as the focal point of this paper.
+The declining trend in Canadian salmon runs, as observed by many avid anglers and supported by empirical data [@citedecline], serves as the focal point of this paper.
 Once thriving waterways have experienced a significant reduction in salmon populations, posing challenges not only for recreational anglers but also for the fishing industry, a vital component of Canadian exports and numerous livelihoods, particularly in coastal regions.
@@
 
-For information regarding the populations of Canadian Salmon in the Pacific region, the *New Salmon Escapement Database System (NuSEDS)* was used to retrieve the quality and health of the Salmon population.
+For information regarding the populations of Canadian Salmon in the Pacific region, the *New Salmon Escapement Database System (NuSEDS)* [@citenuseds] was used to retrieve the quality and health of the Salmon population.
 This data is not only used to measure the populations but other health and success measures that certain area salmon have.
@@
 
-For information regarding the salmon caught in commercial fisheries, the *Pacific Region Commercial Salmon Fishery In-season Catch Estimates* was used.
+For information regarding the salmon caught in commercial fisheries, the *Pacific Region Commercial Salmon Fishery In-season Catch Estimates* [@citecatch] was used.
 This data comes from the Fishery Operations System (FOS) which serves as the primary database for Fisheries and Oceans Canada's (DFO) Pacific Region Commercial Salmon Logbook program, established in 1998 and mandated fleet-wide participation by 2001.

### 2024-04-19 output/paper.qmd
add model math

@@
 
+$$
+\begin{aligned}
+&\mbox{salmon-population}_i|\mu_i, \sigma \sim \mbox{Normal}(\mu_i, \sigma) \\
+\mu_i &= \beta_0 +\beta_1\mbox{vessel-count}_i +\beta_2\mbox{boat-days}_i \\
+& +\beta_3\mbox{salmon-kept}_i +\beta_4\mbox{salmon-released}_i\\
+& +\beta_5\mbox{percentage-gill-net}_i +\beta_6\mbox{percentage-seine}_i\\
+& +\beta_7\mbox{percentage-troll}_i\\
+\beta_{0, 1, ... 7} &\sim \mbox{Normal}(0, 2.5) \\
+\sigma &\sim \mbox{Exponential}(1) \\
+\end{aligned}
+$$
+
 ## Yearly Model
+
+$$
+\begin{aligned}
+&\mbox{next-year-salmon-population}_i|\mu_i, \sigma \sim \mbox{Normal}(\mu_i, \sigma) \\
+\mu_i &= \beta_0 +\beta_1\mbox{adult-broodstock-removals}_i +\beta_2\mbox{jack-broodstock-removals}_i \\
+& +\beta_3\mbox{total-broodstock-removals}_i +\beta_4\mbox{other-removals}_i\\
+& +\beta_5\mbox{total-return-to-river}_i \\
+\beta_{0, 1, ... 5} &\sim \mbox{Normal}(0, 2.5) \\
+\sigma &\sim \mbox{Exponential}(1) \\
+\end{aligned}
+$$
 

### 2024-04-19 output/paper.qmd
update overviews

@@
 abstract: "Using the NuSEDS dataset and Catch Estimates for the Pacific Region Commerical Salmon Fishery dataset provided by the Government of Canada, we investigate the effects of commercial fishing on the reproductive capacity of Canadian Salmon.
-Along with that, we investigate how much of the change in spawn rates can be attributed to the health of the ecosystem, climate, and the fish that live in it.
+Along with that, we can infer how much of the change in spawn rates can be therefore attributed to the health of the ecosystem, climate, and the fish that live in it.
 This paper discusses the impact of direct human intervention in the form of commercial fishing, aiming to highlight the importance in proper management and control in the fishing industry in alignement with changes in the environment by predicting spawning trends of future generations."

### 2024-04-19 scripts/04.2_yearly_change_model.R
update model to relate 4 years later

@@ -22,24 +22,24 @@ broodstock <- read_parquet("output/data/aggregated_broodstock.parquet")
 broodstock_next_year <- 
   broodstock |>
   select(YEAR, TOTAL_RETURN_TO_RIVER) |>
-  mutate(YEAR = YEAR - 1) |>
+  mutate(YEAR = YEAR - 4) |>
   rename(NEXT_YEAR_RETURN = TOTAL_RETURN_TO_RIVER)
 
 # Since for the first and last year, we have nothing to predict, we remove those
 broodstock <- 
   broodstock |>
-  filter(YEAR != max(YEAR))
+  filter(YEAR > min(YEAR) + 4)
 
 broodstock_next_year <-
   broodstock_next_year |>
-  filter(YEAR != min(YEAR))
+  filter(YEAR < max(YEAR) - 4)
 
 # Join the tables (essentially using last years spawning data to predict the next years spawning data)
 model_data <- 
   left_join(broodstock, broodstock_next_year, by = "YEAR") |>
   mutate(
-    TOTAL_RETURN_TO_RIVER = TOTAL_RETURN_TO_RIVER / 1000,
-    NEXT_YEAR_RETURN = NEXT_YEAR_RETURN / 1000
+    TOTAL_RETURN_TO_RIVER = TOTAL_RETURN_TO_RIVER / 10000,
+    NEXT_YEAR_RETURN = NEXT_YEAR_RETURN / 10000
   )
 
 yearly_change_model <- 

### 2024-04-19 output/paper.qmd
finish model section (-modelsummary)

@@
 
-# Data
+# Data {#data}
 
@@
 
+In particular we aim to predict salmon populations based on commercial fishing habits and previous year's spawning health.
+
 ## Catch Model
+
+The first model that was created was intended to discover whether the amount of fish harvested through commercial fisheries and the amount of vessels actively harvesting significantly impacted salmon population in Pacific Canada.
+
+We expect that an increase in the amount of fish harvested would impact the population, so we aim to ensure that commercial fishing was and will be done in a controlled and sustainable way.
+Here we are really testing the effectiveness of fishing regulations and license allocations that the Canadian government has put in place to protect salmon populations.
+
+Therefore the model we are interested in is:
 
@@
 
+`salmon-population`, `vessel-count`, `boat-days`, `salmon-kept` and `salmon-released` are all as they were in @tbl-catchdata_gill, @tbl-catchdata_seine, and @tbl-catchdata_troll, accounting for the frequency and quantity of harvest.
+
+`percentage-gill-net`, `seine`, and `troll` account for how much of the total amount of salmon harvested came from that respective method, since all salmon species and fishing methods were aggregated as mentioned in the [data](#data) section.
+
+These variables are included to see if certain methods of fishing have a greater or lesser effect on the salmon population.
+
+```{r}
+# TODO modelsummary
+```
+
 ## Yearly Model
+
+The second model uses past year's spawning data to predict the next year's populations.
+It is reasonable to expect that bad spawning rates one year can affect future year's spawning rates.
+With NOAA stating that the most common age of returning and spawning adults being 4 years old [@citelifespan], the predicted variable is the salmon populations observed 4 years in the future.
+
+There are also 2 types of spawning data as shown in @tbl-broodstock vs @tbl-spawnpopulation, being natural vs human cultivated, so the model takes into account those factors, aiming to see the the natural sustainability of salmon populations as well as the impact of protected and selective breeding on those populations.
+
+Therefore the model we are interested in is:
 
@@
 $$
+
+Where the variables are defined the same way as mentioned when discussing @tbl-broodstock and @tbl-spawnpopulation, with the except to `next-year-salmon-population`, which is the population of salmon 4 years after a certain observation was made as mentioned above.
+
+```{r}
+# TODO modelsummary
+```
 

### 2024-04-19 output/paper.qmd
add modelsummaries

@@
 
-To create the linear model we made use of the `rstanarm` [@citerstanarm] package.
+To create the linear model we made use of the `rstanarm` [@citerstanarm] package, and to summarize the results we made use of the `modelsummary` [@citemodelsummary] package.
 
 In particular we aim to predict salmon populations based on commercial fishing habits and previous year's spawning health.
+
+```{r}
+#| include: false
+#| warning: false
+#| message: false
+#| echo: false
+
+# Load in models
+yearly_change_model <-
+  readRDS(file = "models/yearly_change_model.rds")
+
+catch_population_model <-
+  readRDS(file = "models/catch_population_model.rds")
+
+```
 
@@
 
+@tbl-catchmodel shows a model summary for this model, this will be discussed further in the [results](#results) section.
+
 ```{r}
-# TODO modelsummary
+#| message: false
+#| warning: false
+#| echo: false
+#| label: tbl-catchmodel
+#| tbl-cap: "Predicting the Salmon Population given commercial fishing catch estimates"
+
+modelsummary(
+  list(
+    "Salmon Population" = catch_population_model
+  ),
+  statistic = "mad"
+  )
+
 ```
@@
 
+@tbl-yearlymodel shows a model summary for this model, this will be discussed in the [results](#results) section.
+
 ```{r}
-# TODO modelsummary
+#| message: false
+#| warning: false
+#| echo: false
+#| label: tbl-yearlymodel
+#| tbl-cap: "Predicting the Salmon Population in 4 years time given a year's spawn statistics"
+
+modelsummary(
+  list(
+    "Salmon Population" = yearly_change_model
+  ),
+  statistic = "mad"
+  )
+
 ```
 
-# Results
+# Results {#results}
 

### 2024-04-19 output/paper.qmd
complete model + data sections

@@
 ```{r}
-# TODO show all the relationships (broodstock vs population, total catches vs population, type vs population)
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-cap: "Salmon Catch Data from 2002-2022 (Gill)"
+#| fig-subcap: ["Coho and Chinook", "Chum, Pink, and Sockeye"]
+#| label: fig-catchgraph_gill
+#| layout-ncol: 2
+
+pivoted_catch_data <- 
+  gill_net_catches |>
+  select(-STEELHEAD_KEPT, -SOCKEYE_KEPT, -CHUM_KEPT, -PINK_KEPT) |>
+  pivot_longer(cols = contains("_KEPT"), names_to = "Species", values_to = "Count")
+
+ggplot(pivoted_catch_data, aes(x = CALENDAR_YEAR, y = Count, fill = Species)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Calendar Year", y = "Count", fill = "Species") +
+  theme_minimal()
+
+pivoted_catch_data <- 
+  gill_net_catches |>
+  select(-STEELHEAD_KEPT, -CHINOOK_KEPT, -COHO_KEPT) |>
+  pivot_longer(cols = contains("_KEPT"), names_to = "Species", values_to = "Count")
+
+ggplot(pivoted_catch_data, aes(x = CALENDAR_YEAR, y = Count, fill = Species)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Calendar Year", y = "Count", fill = "Species") +
+  theme_minimal()
+
 ```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-cap: "Salmon Catch Data from 2002-2022 (Seine)"
+#| fig-subcap: ["Coho and Chinook", "Chum, Pink, and Sockeye"]
+#| label: fig-catchgraph_seine
+#| layout-ncol: 2
+
+pivoted_catch_data <- 
+  seine_catches |>
+  select(-STEELHEAD_KEPT, -SOCKEYE_KEPT, -CHUM_KEPT, -PINK_KEPT) |>
+  pivot_longer(cols = contains("_KEPT"), names_to = "Species", values_to = "Count")
+
+ggplot(pivoted_catch_data, aes(x = CALENDAR_YEAR, y = Count, fill = Species)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Calendar Year", y = "Salmon Kept", fill = "Species") +
+  theme_minimal()
+
+pivoted_catch_data <- 
+  seine_catches |>
+  select(-STEELHEAD_KEPT, -CHINOOK_KEPT, -COHO_KEPT) |>
+  pivot_longer(cols = contains("_KEPT"), names_to = "Species", values_to = "Count")
+
+ggplot(pivoted_catch_data, aes(x = CALENDAR_YEAR, y = Count, fill = Species)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Calendar Year", y = "Salmon Kept", fill = "Species") +
+  theme_minimal()
+
+```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-cap: "Salmon Catch Data from 2002-2022 (Troll)"
+#| fig-subcap: ["Coho and Chinook", "Chum, Pink, and Sockeye"]
+#| label: fig-catchgraph_troll
+#| layout-ncol: 2
+
+pivoted_catch_data <- 
+  troll_catches |>
+  select(-STEELHEAD_KEPT, -SOCKEYE_KEPT, -CHUM_KEPT, -PINK_KEPT) |>
+  pivot_longer(cols = contains("_KEPT"), names_to = "Species", values_to = "Count")
+
+ggplot(pivoted_catch_data, aes(x = CALENDAR_YEAR, y = Count, fill = Species)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Calendar Year", y = "Salmon Kept", fill = "Species") +
+  theme_minimal()
+
+pivoted_catch_data <- 
+  troll_catches |>
+  select(-STEELHEAD_KEPT, -CHINOOK_KEPT, -COHO_KEPT) |>
+  pivot_longer(cols = contains("_KEPT"), names_to = "Species", values_to = "Count")
+
+ggplot(pivoted_catch_data, aes(x = CALENDAR_YEAR, y = Count, fill = Species)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Calendar Year", y = "Salmon Kept", fill = "Species") +
+  theme_minimal()
+
+```
+
+@fig-catchgraph_gill, @fig-catchgraph_seine, and @fig-catchgraph_troll show comparitive catch estimates between the different methods of fishing with some years finding more use or success in certain methods over others.
+With these variations, we aim to analyze if certain methods should be used over others due to their effectiveness, sustainability, or otherwise.
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-cap: "Salmon Broodstock Removals From 1975 - 2022"
+#| label: fig-broodstockgraph
+
+pivoted_broodstock_data <- 
+  broodstock |>
+  filter(YEAR >= 1975) |>
+  select(YEAR, ADULT_BROODSTOCK_REMOVALS, JACK_BROODSTOCK_REMOVALS) |>
+  pivot_longer(cols = c(ADULT_BROODSTOCK_REMOVALS, JACK_BROODSTOCK_REMOVALS), names_to = "Removal_Type", values_to = "Count")
+
+ggplot(pivoted_broodstock_data, aes(x = YEAR, y = Count, fill = Removal_Type)) +
+  geom_bar(stat = "identity", position = "stack") +
+  labs(x = "Year", y = "Broodstock Removed", fill = "Removal Type") +
+  theme_minimal()
+
+```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-cap: "Salmon Populations From 1975 - 2022"
+#| label: fig-populationgraph
+
+broodstock |>
+  filter(YEAR >= 1975) |>
+  ggplot(aes(x = YEAR, y = TOTAL_RETURN_TO_RIVER)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Year", y = "Total Return to River") +
+  theme_minimal()
+```
+
+@fig-broodstockgraph and @fig-populationgraph show the trends of population in relation to breeding efforts.
+It is evident that as salmon populations have declined, more effort is being made in trying to repopulate as shown by the increase in broodstock removals in @fig-broodstockgraph.
+With this, the model will aim to capture the true effectiveness of these efforts.
+
+For the purposes of this report, salmon species were grouped together and total populations irrespective of area were considered instead.
+The health of salmon populations within species is closely connected, spawning at the same time, maturing roughly at around the same time [@citelifespan].
+Due to this reasoning as well as testing with separate species showing similar results it was decided that species would be considered all the same for the purposes of this report.
 

### 2024-04-19 output/paper.qmd
complete paper

@@
 abstract: "Using the NuSEDS dataset and Catch Estimates for the Pacific Region Commerical Salmon Fishery dataset provided by the Government of Canada, we investigate the effects of commercial fishing on the reproductive capacity of Canadian Salmon.
-Along with that, we can infer how much of the change in spawn rates can be therefore attributed to the health of the ecosystem, climate, and the fish that live in it.
+Along with that, we can infer how much of the change in spawn rates can affect future populations, and whether stocking programs have been able to mitigate population decline.
 This paper discusses the impact of direct human intervention in the form of commercial fishing, aiming to highlight the importance in proper management and control in the fishing industry in alignement with changes in the environment by predicting spawning trends of future generations."
@@
 
+## Catch Model
+
+The model summary from @tbl-catchmodel shows the relationship between fishing activity and salmon populations.
+The model suggests that an increase in the amount of commercial vessels fishing for salmon on the water correlates with a decrease in the salmon populations.
+This is an expected result, with more pressure on fish, the more the population will suffer.
+Surprisingly however, the number of days spent fishing shows a slightly positive correlation.
+This unexpected finding suggests that spreading out fishing activity over more days may alleviate pressure on salmon populations, contributing to their preservation.
+
+Then looking to the actual catch numbers, it seems that the actual amount of salmon caught doesn't have a major impact on the salmon populations, which is a good result to have.
+This means that the restrictive measures placed on commercial fishers are working and are preventing the fishing industry from irreversibly depleting this resource.
+This suggests that it is not just about the amount of salmon that are actually removed from the ecosystem, but rather how much human interference from the mere act of fishing an ecosystem can handle.
+Fishing for salmon (and any other fish for that matter) in a destructive way can have impacts on the food chain that affect the salmon the same, even if the catch numbers do not portray that fact.
+
+More interestingly, there is a significant difference between the different methods of fishing to the salmon population.
+Specifically, @tbl-catchmodel suggests that when trolling is the dominant fishing method, salmon populations tend to be lower compared to periods when gill nets are prevalent.
+Seine nets, on the other hand, represent a middle ground, with lesser effects on salmon populations.
+These findings show the importance of considering the ecological implications of different salmon harvesting methods and the need for targeted management strategies to sustainably manage future salmon populations.
+
+## Yearly Model
+
+From the model shown in summary by @tbl-yearlymodel, the removal of salmon for broodstock only shows a slightly negative or possibly neutral relationship with future salmon population.
+The human influence through breeding and stocking salmon does not seem to have a greatly positive or negative effect on the overall population, but it suggests that other factors are at play.
+As mentioned in the [data](#data) section and shown in @fig-broodstockgraph @fig-populationgraph, it seems that while broodstock removals increase, population decreases, and since this model doesn't suggest a strong correlation between them, they must be connected in some other way.
+It is reasonable to say that this actually suggests that increased broodstock removals are a response to declining population, and are aimed at restoring losses that were occurred another way.
+While increased breeding doesn't increase the population over what it was before, it maintains it, appearing as a neutral relationship in a linear model.
+To be truly sure, research into the breeding and stocking process must be done to verify these results.
+
+Looking at the return to river variable in @tbl-yearlymodel, this suggests that most of the increase in population is closely related to how succesfully and plentiful the spawning was 4 years ago.
+The smolt (baby salmon), that hatched 4 years ago, most likely become the salmon that appear in the waterways to spawn now after they have reached maturity.
+This shows how important future planning is in salmon conservation, as impacts on population may not be seen immediately.
+Preventative measures are necessary, as reactive measures to suddenly declining salmon populations will not take effect for a significant period of time.
+
 # Discussion {#discussion}
 
-```{r}
-# TODO talk about the fact that region caught vs population count region were not considered,
-# to be more accuracte, we would monitor changes between each watershed
+## Weaknesses
 
-# Talk about the large amount of na values, population may not be accurately represented in some years
-```
+The datasets that were used provided location data in the forms of management regions which were not made use of in this dataset.
+Especially for the spawning regions, other factors could be taken into account to get a better understanding on what hardships a region is facing, and what fishing regions have greater impact to those spawning regions.
+This paper dealt with the general impacts and changes of Pacific Canadian salmon, but working at a smaller scale may help with identifying regional differences and allow different approaches depending on such.
+
+In addition, enviornmental factors were not considered in this report, although they were suggested to have correlation, due gaps in explanation for change just from human intervention.
+Disease, climate, food chain fluctuations, can all have impacts on the health and abundance of the salmon population, factors that can be changed in other ways and through different regulations.
+
+As for the data itself, there were a lot of missing values, which may impact the accuracy of the results, with some waterbodies / watersheds potentially being underrepresented when calculating population.
+Similarly, both the salmon population and catch data are both estimates made by Fisheries and Oceans Canada, meaning that the counts are not 100% accurate.
+If these estimates were made erroneously, this may also impact the accuracy of the models and results.
+
+## What was done and should be done
+
+Through the use of linear models, we were able to make inferences on the world, in this case the world of salmon fisheries, results that could have impacts for both recreational anglers and the commercial fishing industry.
+By analyzing past fishing trends and spawning statistics, we were able to verify our expecations about human driven impacts on salmon populations.
+We saw the successes and failings of restrictive fishing regulations as well as how salmon breeding and stocking has supported the salmon population during a period of decline.
+
+From this analysis of salmon's decline, we aim to reiterate the importance of mitigation strategies and management plans to restore salmon populations and promote sustainable fishing practices.
+Governing bodies should focus on what has been working and put further emphasis in places which have been failing.
+Preserving healthy salmon populations is not only vital for biodiversity and ecosystem balance but also for meeting the demands of the fishing industry.
+Despite the significant correlations found between fishing frequency, previous salmon populations, and artificial breeding practices, it is clear that other factors not captured in the available data may also contribute to salmon population fluctuations.
+Further investigation into these factors should be done to gain a more comprehensive understanding of the dynamics affecting salmon populations and to inform effective conservation efforts moving forward.
 

### 2024-04-19 scripts/00_simulation.R
create simulations

@@ -1,0 +1,53 @@
+#### Preamble ####
+# Purpose: Simulates data for predicting salmon populations given commercial fishing and salmon stocking and spawning
+# Author: Timothius Prajogi
+# Data: April 18 2024
+# Contact: timothius.prajogi@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+
+#### Workspace setup ####
+
+library(tidyverse)
+set.seed(302)
+
+#### Simulate data ####
+
+# Simulate a negative relation between salmon caught commercially 
+# and the overall salmon population
+salmon_population <- sample(100:1000, 18, replace = TRUE)
+salmon_caught <- salmon_population + rnorm(18, mean = -50, sd = 50)
+salmon_population <- salmon_population - salmon_caught
+
+# simulate results from 2005 to 2022
+data <- data.frame(
+  year = 2005:2022,
+  salmon_population = salmon_population,
+  salmon_caught = salmon_caught
+)
+
+ggplot(data, aes(x = year)) +
+  geom_line(aes(y = salmon_population, color = "Salmon Population")) +
+  geom_line(aes(y = salmon_caught, color = "Salmon Caught")) +
+  labs(x = "Year", y = "Count", title = "Salmon Population vs. Salmon Caught") +
+  scale_color_manual(values = c("blue", "red"), labels = c("Salmon Population", "Salmon Caught")) +
+  theme_minimal()
+
+# simulate that an increase in population and an increase in the amount of
+# salmon artificially breed then released increases future populations
+salmon_population <- seq(100, 1000, length.out = 18) + rnorm(18, mean = 0, sd = 50)
+salmon_stocking <- rnorm(18, mean = 50, sd = 50)
+salmon_population <- salmon_population + salmon_stocking
+
+sim_data <- data.frame(
+  year = 2005:2022,
+  salmon_population = salmon_population,
+  salmon_stocking = salmon_stocking
+)
+
+ggplot(sim_data, aes(x = year)) +
+  geom_line(aes(y = salmon_population, color = "Salmon Population")) +
+  geom_line(aes(y = salmon_stocking, color = "Salmon Stocking")) +
+  labs(x = "Year", y = "Count", title = "Salmon Population vs. Salmon Stocking") +
+  scale_color_manual(values = c("blue", "red"), labels = c("Salmon Population", "Salmon Stocking")) +
+  theme_minimal()

### 2024-04-19 scripts/03.1_aggregate_catch_data.R
remove duplicate code

@@ -149,39 +149,4 @@ pivoted_catch_data <-
 
 write_parquet(pivoted_catch_data, "output/data/aggregated_pivoted_catch_data.parquet")
 
-# Aggregate spawning population data
-
-spawning_population <- read_parquet('output/data/spawning_population.parquet')
-
-# Group first by the species, area, and year
-aggregated_spawning_population_by_area <- 
-  spawning_population |>
-  mutate(SALMON_POPULATION = coalesce(TOTAL_RETURN_TO_RIVER, NATURAL_ADULT_SPAWNERS)) |>
-  filter(SPECIES != "Atlantic" & SPECIES != "Kokanee") |>
-  group_by(SPECIES, AREA, YEAR) |>
-  summarize(
-    NATURAL_ADULT_SPAWNERS = sum(NATURAL_ADULT_SPAWNERS, na.rm = TRUE),
-    NATURAL_JACK_SPAWNERS = sum(NATURAL_JACK_SPAWNERS, na.rm = TRUE),
-    NATURAL_SPAWNERS_TOTAL = sum(NATURAL_SPAWNERS_TOTAL, na.rm = TRUE),
-    TOTAL_RETURN_TO_RIVER = sum(TOTAL_RETURN_TO_RIVER, na.rm = TRUE),
-    SALMON_POPULATION = sum(SALMON_POPULATION, na.rm = TRUE)
-  )
-
-# Rename values to be capitalized
-aggregated_spawning_population_by_area$SPECIES = toupper(aggregated_spawning_population_by_area$SPECIES)
-
-# Group together only by species and year
-aggregated_spawning_population <- 
-  aggregated_spawning_population_by_area |>
-  group_by(SPECIES, YEAR) |>
-  summarize(
-    NATURAL_ADULT_SPAWNERS = sum(NATURAL_ADULT_SPAWNERS, na.rm = TRUE),
-    NATURAL_JACK_SPAWNERS = sum(NATURAL_JACK_SPAWNERS, na.rm = TRUE),
-    NATURAL_SPAWNERS_TOTAL = sum(NATURAL_SPAWNERS_TOTAL, na.rm = TRUE),
-    TOTAL_RETURN_TO_RIVER = sum(TOTAL_RETURN_TO_RIVER, na.rm = TRUE),
-    SALMON_POPULATION = sum(SALMON_POPULATION, na.rm = TRUE)
-  )
-
-write_parquet(aggregated_spawning_population, "output/data/aggregated_spawning_population.parquet")
-  
   

### 2024-04-19 output/paper.qmd
cite testthat

@@
 
-For additional assistance with data storage, the `arrow` [@citearrow] was used and for report generation, the `knitr` [@citeknitr] package was used.
+For additional assistance with data storage, the `arrow` [@citearrow] package was used and for report generation, the `knitr` [@citeknitr] package was used.
+For testing / verifying data the `testthat` [@citetestthat] package was used.
 

### 2024-04-19 scripts/03.2_aggregated_spawn_data.R
convert to ints

@@ -48,6 +48,12 @@ aggregated_spawning_population <-
     SALMON_POPULATION = sum(SALMON_POPULATION, na.rm = TRUE)
   )
 
+aggregated_spawning_population$NATURAL_ADULT_SPAWNERS <- 
+  round(aggregated_spawning_population$NATURAL_ADULT_SPAWNERS)
+
+aggregated_spawning_population$SALMON_POPULATION <- 
+  round(aggregated_spawning_population$SALMON_POPULATION)
+
 write_parquet(aggregated_spawning_population, "output/data/aggregated_spawning_population.parquet")
 
 # Aggregate broodstock data
@@ -64,5 +70,8 @@ aggregated_broodstock <-
   ) |>
   filter(YEAR >= 1979) |>
   arrange(YEAR)
+
+aggregated_broodstock$TOTAL_RETURN_TO_RIVER <- 
+  round(aggregated_broodstock$TOTAL_RETURN_TO_RIVER)
 
 write_parquet(aggregated_broodstock, "output/data/aggregated_broodstock.parquet")

### 2024-04-19 scripts/05_test_data.R
create data verification script

@@ -1,0 +1,79 @@
+#### Preamble ####
+# Purpose: Validate data
+# Author: Timothius Prajogi
+# Date: 18 April 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01_download_data, 02_clean_data, 03_aggregate_data
+
+#### Workplace setup ####
+
+library(testthat)
+library(tidyverse)
+
+#### Validate Data ####
+
+catch_data <- read_parquet("output/data/aggregated_pivoted_catch_data.parquet")
+spawning_data <- read_parquet("output/data/aggregated_spawning_population.parquet")
+broodstock_data <- read_parquet("output/data/aggregated_broodstock.parquet")
+
+# Catch data verification
+
+# Check if years are in the range 2005 to 2022 inclusive
+expect_true(all(catch_data$CALENDAR_YEAR >= 2005 & catch_data$CALENDAR_YEAR <= 2022),
+            info = "Years should be in the range 2005 to 2022 inclusive")
+
+# Check if VESSEL_COUNT and BOAT_DAYS are integers
+expect_true(all(catch_data$VESSEL_COUNT %% 1 == 0),
+            info = "VESSEL_COUNT should be integers")
+expect_true(all(catch_data$BOAT_DAYS %% 1 == 0),
+            info = "BOAT_DAYS should be integers")
+
+# Check if all kept and released values are integers
+expect_true(all(catch_data$SALMON_KPT %% 1 == 0),
+            info = "SALMON_KEPT should contain integers")
+expect_true(all(catch_data$SALMON_RLD %% 1 == 0),
+            info = "SALMON_RLD should contain integers")
+
+# Check that the species are in the expected set
+allowed_species <- c("SOCKEYE", "COHO", "PINK", "CHUM", "CHINOOK", "STEELHEAD")
+expect_true(all(catch_data$SPECIES %in% allowed_species),
+            info = "SPECIES column should contain only specific species")
+
+# Spawning data verification
+
+expect_true(all(unique(spawning_data$SPECIES) %in% allowed_species),
+            info = "SPECIES column should contain only specific species")
+
+# Check if YEAR column values are within the specified range
+expect_true(all(spawning_data$YEAR >= 1950 & spawning_data$YEAR <= 2022),
[…]

### 2024-04-19 scripts/05_test_data.R
Update 05_test_data.R

@@ -67,13 +67,17 @@ expect_true(all(spawning_data$TOTAL_RETURN_TO_RIVER %% 1 == 0),
 expect_true(all(spawning_data$SALMON_POPULATION %% 1 == 0),
             info = "SALMON_POPULATION column values should be integers")
 
-expect_true(all(catch_data$ADULT_BROODSTOCK_REMOVALS %% 1 == 0), 
-            msg = "ADULT_BROODSTOCK_REMOVALS column contains only integers.")
-expect_true(all(catch_data$JACK_BROODSTOCK_REMOVALS %% 1 == 0), 
-            msg = "JACK_BROODSTOCK_REMOVALS column contains only integers.")
-expect_true(all(catch_data$TOTAL_BROODSTOCK_REMOVALS %% 1 == 0), 
-            msg = "TOTAL_BROODSTOCK_REMOVALS column contains only integers.")
-expect_true(all(catch_data$OTHER_REMOVALS %% 1 == 0), 
-            msg = "OTHER_REMOVALS column contains only integers.")
-expect_true(all(catch_data$TOTAL_RETURN_TO_RIVER %% 1 == 0), 
-            msg = "TOTAL_RETURN_TO_RIVER column contains only integers.")
+expect_true(all(broodstock_data$ADULT_BROODSTOCK_REMOVALS %% 1 == 0), 
+            info = "ADULT_BROODSTOCK_REMOVALS column contains only integers.")
+
+expect_true(all(broodstock_data$JACK_BROODSTOCK_REMOVALS %% 1 == 0), 
+            info = "JACK_BROODSTOCK_REMOVALS column contains only integers.")
+
+expect_true(all(broodstock_data$TOTAL_BROODSTOCK_REMOVALS %% 1 == 0), 
+            info = "TOTAL_BROODSTOCK_REMOVALS column contains only integers.")
+
+expect_true(all(broodstock_data$OTHER_REMOVALS %% 1 == 0), 
+            info = "OTHER_REMOVALS column contains only integers.")
+
+expect_true(all(broodstock_data$TOTAL_RETURN_TO_RIVER %% 1 == 0), 
+            info = "TOTAL_RETURN_TO_RIVER column contains only integers.")

### 2024-04-19 other/datasheets/datasheet_catch.qmd
complete datasheets

---
title: "Datasheet for 'Pacific Region Commercial Salmon Fishery In-season Catch Estimates'"
author: 
  - Timothius Prajogi
date: today
date-format: long
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The database was built in 2001 and continues to be the current repository for all data pertaining to DFO’s Pacific Region Commercial Salmon Logbook program.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - Fisheries and Oceans Canada
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - Government funded.
4. *Any other comments?*
    - No.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- The instances of the dataset represent catch estimates for all types of salmon, the amount of vessels that were licensed, the amount of days that were fished by boats in total, and the year and area this took place in. They are separated by the type of fishing that was done.
2. *How many instances are there in total (of each type, if appropriate)?*
	- 238 observations for gill nets, 190 for seine nets, 468 for trolling vessels.
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- It contains all recoreded observations. 
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- It consists of total count values of each feature.
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- The rows are labeled by the fishery and licence area.
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- Some observations have been redacted due to privacy reasons.
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- No.
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- No.
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- No.
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- Self contained.
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- Confidential data has been removed.
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- No.
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- It identifies salmon by their genus/species.
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- No.
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- No.
16. *Any other comments?*
	- No.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- Fishers were instructed to log their catches, and total estimates were made by Fisheries and Oceans Canada based on those logs.
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- Human curation with the assistance of programs helped create these estimates.
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- No.
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- Both the fishers and the researchers were involved in creating this dataset.
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- From 2005 - 2022.
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Not stated.
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- I collected directly from the source.
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- No.
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- No they were not, they had licensed the dataset to not require such authorization.
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- N/A
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Not stated.
12. *Any other comments?*
	- No.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- The rows ommitted for privacy reasons were removed and skipped.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- Yes, the raw data in this project can be found in `input/data` and can be found from the source here: https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6.
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- Yes, R a statistical programming language was used and the documentation can be found here: https://www.r-project.org/
4. *Any other comments?*
	- No.

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- Not stated.
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- No.
3. *What (other) tasks could the dataset be used for?*
	- The dataset could be used to monitor the fishing industry, its success and failures as well as its impact on other ecosystems and environments.
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- No.
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- It is to be noted that this dataset is only for the Canadian Pacific, and should only be used in analyses that are about that region.
6. *Any other comments?*
	- No.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- Yes, it is distributed as part of Canada's open data portal.
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- It will be downloadable through a provided API endpoint.
3. *When will the dataset be distributed?*
	- It is updated quarterly and was published in 2016
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- It is published under the Open Government Licence - Canada
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- Not stated.
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- Not stated.
7. *Any other comments?*
	- No.

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- Maintenance is done by the publisher Fisheries and Oceans Canada.
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- They have provided an email address to be contacted: shelee.hamilton@dfo-mpo.gc.ca
3. *Is there an erratum? If so, please provide a link or other access point.*
	- No.
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- Not stated.
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- N/A
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- Not stated.
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- Not stated.
8. *Any other comments?*
	- No.

\newpage

# References

### 2024-04-19 other/datasheets/datasheet_nuseds.qmd
complete datasheets

--- 2024-04-19 other/datasheets/datasheet_catch.qmd
@@
 ---
-title: "Datasheet for 'Pacific Region Commercial Salmon Fishery In-season Catch Estimates'"
+title: "Datasheet for 'NuSEDS-New Salmon Escapement Database System'"
 author: 
@@
 1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
-    - The database was built in 2001 and continues to be the current repository for all data pertaining to DFO’s Pacific Region Commercial Salmon Logbook program.
+    - The dataset was created to keep track and monitor salmon spawning records and spawner abundance estimates.
+It is the primary database used to track salmon for Fisheries and Oceans Canada and it enables future research.
+It reports spawning observations for 9100_ individual populations.
 2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
-    - Fisheries and Oceans Canada
+    - The dataset was published by Fisheries and Oceans Canada.
 3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
-    - Government funded.
+    - This is a government led program.
 4. *Any other comments?*
@@
 Please provide a description.*
-	- The instances of the dataset represent catch estimates for all types of salmon, the amount of vessels that were licensed, the amount of days that were fished by boats in total, and the year and area this took place in.
-They are separated by the type of fishing that was done.
+	- An instance in a dataset represents the spawning salmon population observed in a specific waterbody during a specific year.
 2. *How many instances are there in total (of each type, if appropriate)?*
-	- 238 observations for gill nets, 190 for seine nets, 468 for trolling vessels.
+	- There are 413,434 observations made.
 3.
@@
 If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
-	- It contains all recoreded observations. 
+	- The dataset contains all recorded observations made in the Pacific waterbodies of Canada, and is intended to be a comprehensive dataset.
 4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
-	- It consists of total count values of each feature.
+	- Each observation notes down amount of salmon observed, when it was observed, the type of salmon, any broodstock removed, and the waterbody the observation was made in. 
 5. *Is there a label or target associated with each instance? If so, please provide a description.*
-	- The rows are labeled by the fishery and licence area.
+	- Each observation is labeled with a unique identifier labeled `ACT_ID`.
 6.
@@
 This does not include intentionally removed information, but might include, for example, redacted text.*
-	- Some observations have been redacted due to privacy reasons.
+	- The instance may or may not record certain data regarding the spawning population.
+For broodstock data, when it is not present it signifies no broodstock was removed from the waterway.
+Specific date values are sometimes missing, being insignificant data as spawning season generally comes around at the same time, with specific dates recorded only for certain waterbodies.
 7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
-	- No.
+	- The waterbodies and regions are combined to create a unique id which can be used to reference all other observations made about that waterbody and region for any year, species, or otherwise.
 8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
-	- No.
+	- No datasplits were recommended.
 9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
-	- No.
+	- Historical data that was imported was included but is largely blank. This data just recorded basic observations, with newer observations provided much more detail.
 10.
@@
 Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
-	- Self contained.
+	- The dataset is self-contained, hosted on Canada's open data portal.
 11.
@@
 If so, please provide a description.*
-	- Confidential data has been removed.
+	- No.
 12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
@@
 If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
-	- It identifies salmon by their genus/species.
+	- As mentioned the observations were split by species, waterbody, and region. These attributes are identified in their respective columns.
 14.
@@
 If so, please describe how.*
-	- Fishers were instructed to log their catches, and total estimates were made by Fisheries and Oceans Canada based on those logs.
+	- The data was collected as a survey, made from observations gathered from field researchers.
+Estimates were made as to the exact counts, with the scale of the matter being so large and spanning long periods.
 2.
@@
 How were these mechanisms or procedures validated?*
-	- Human curation with the assistance of programs helped create these estimates.
+	- Human curation was necessary to gather the data, along with measurement instruments.
 3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
-	- No.
+	- N/A
 4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
-	- Both the fishers and the researchers were involved in creating this dataset.
+	- This is not explicitly mentioned, however the data was provided by Fisheries and Oceans Canada.
 5.
@@
 If not, please describe the timeframe in which the data associated with the instances was created.*
-	- From 2005 - 2022.
+	- The data is collected yearly, and has been collected since 1950 with the last update being 2022. The dataset is updated quarterly as finalized observations come in.
 6.
@@
 If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
-	- Not stated.
+	- No ethical review processes were stated.
 7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
-	- I collected directly from the source.
+	- I collected the data directly from their API source.
 8.
@@
 If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
-	- No.
+	- No they were not, they had licensed the dataset to not require such authorization.
 9.
@@
 If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
-	- No they were not, they had licensed the dataset to not require such authorization.
+	- N/A
 10.
@@
 If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
-	- Not stated.
+	- No analysis was stated.
 12. *Any other comments?*
-	- No.
+	- The background review process for this dataset was not stated, being a government database.
 
@@
 If not, you may skip the remaining questions in this section.*
-	- The rows ommitted for privacy reasons were removed and skipped.
+	- Yes, N/A values needed to be handled, either treated as 0, or their observations ignored.
+Depending on the requirements, salmon species were grouped together by year to get an understanding of the grander scale of the population.
 2.
@@
 1. *Has the dataset been used for any tasks already? If so, please provide a description.*
-	- Not stated.
+	- I imagine so, but usage has not been documented.
 2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
@@
 3. *What (other) tasks could the dataset be used for?*
-	- The dataset could be used to monitor the fishing industry, its success and failures as well as its impact on other ecosystems and environments.
+	- The dataset covers a lot of features that can be used in various aspects of salmon health monitoring.
+Other research surrounding each of the different features, spawn timings, broodstock, population, can be done.
 4.
@@
 2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
-	- They have provided an email address to be contacted: shelee.hamilton@dfo-mpo.gc.ca
+	- They have provided an email address to be contacted: Zhipeng.Wu@dfo-mpo.gc.ca
 3. *Is there an erratum? If so, please provide a link or other access point.*
